\chapter{Swapping mechanisms}
\label{ch:swapping-mechanisms}

\begin{reading}
  \begin{enumerate}
  \item
    \citet[Chapter 21]{arpaci-dusseau:2015:operating}\\
    \url{http://pages.cs.wisc.edu/~remzi/OSTEP/vm-beyondphys.pdf}
  \end{enumerate}
\end{reading}

\section{Assumptions thus far}

\begin{enumerate}
\item Process address space small compared to physical memory.
\item All running process' address spaces fit into physical memory.
\end{enumerate}

\section{Swap space}

Usually dedicated area on disk (slower, cheaper than RAM).

For now, assume very large.

Consider effect of loading running binary from disk (not necesarily swap space, but same idea).

\section{Page table}

\begin{description}
\item[Present bit] is 1 if a VPN is in a PFN in physical memory. 0 otherwise.
\item[Page fault] said to occur if PTE for VPN has 0 for present bit.
\item[Page fault handler] runs to handle a page fault when it occurs (part of OS).
\end{description}

\section{Handling page fault}

\begin{enumerate}
\item Process blocked while page fault handled.
\item Find page on disk (may use PTE bits for PFN to store disk address).
\item Must swap page into memory.
\item Update page table: mark as present, update PFN.
\item Retry instruction:
  \begin{itemize}
  \item May generate TLB miss, which is handled as normal.
  \item Or may update TLB when handling page fault.
  \end{itemize}
\end{enumerate}

\section{Paging operations}

\begin{description}
\item[Page in] where PFN loaded from disk.
\item[Page out] where PFN written to disk.
\item[Page replacement policy] decides what PFN to page out when memory is full.
\end{description}

\section{Control flow}

Possibilities:

\begin{enumerate}
\item Page present and valid: lookup PTE, cache in TLB.
\item Page not present but valid: page in, update PTE, lookup.
\item Page invalid: trap to OS (probably terminate process)
\end{enumerate}

\section{Swap daemon}

\begin{description}
\item[Low watermark] is lowest number of free pages we'd like.
\item[High watermark] is enough free pages.
\end{description}

Once OS sees fewer than LW pages free, swap/page daemon pages out PFNs to disk until HW pages area available.

(Will generally do this in groups, like I/O.)



%%% Local Variables:
%%% mode: latex
%%% TeX-master: "os_notes"
%%% End:

